\documentclass[10pt,a4paper]{article}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fullpage}
\usepackage{color}

\author{Patrick Lam}

\parindent=0in
\newcommand{\red}{\color{red}}
\newcommand{\black}{\color{black}}
\begin{document}

\begin{center}
\begin{Large}Week 4 Problems\end{Large}
\end{center}
\bigskip
\begin{enumerate}

\item (Adapted from Gelman 2.10) Suppose there are $N$ cable cars in San Francisco, numbered sequentially from 1 to $N$.  You see a cable car at random; it is numbered 203.  You wish to estimate $N$. \\

Assume your prior distribution on $N$ is geometric with mean 100; that is,
\begin{eqnarray*}
p(N) = (1/100)(99/100)^{N-1}, \; \mathrm{for} \; N = 1,2,\dots
\end{eqnarray*}

\begin{itemize}
\item[a)] What is your posterior distribution for $N$ up to a constant of proportionality?   
\item[b)] Find the posterior mean and standard deviation by approximating the normalizing constant in R (without simulating). 
\item[c)] Now find the posterior mean and standard deviation by simulation in R.  Are your answers similar to those in b)?
\end{itemize}

\bigskip

\item (adapted from Gelman 3.2) On September 25, 1988, the evening of a Presidential campaign debate, ABC News conducted a survey of registered voters in the United States; 639 persons were polled before the debate, and 639 different persons were polled after.  \\

\begin{table}[!htp]
\begin{center}
\begin{tabular}{ccccc}
Survey & Bush & Dukakis & No opinion/other & Total \\
\hline
pre-debate & 294 & 307 & 38 & 639 \\
post-debate & 288 & 332 & 19 & 639 \\
\end{tabular}
\end{center}
\end{table}

Assume the surveys are independent simple random samples from the population of registered voters.  Model the data with two different multinomial distributions.
\begin{enumerate}
\item[a)] What is the posterior probability that support for Bush increased between the two surveys?
\item[b)] Of the voters who had a preference for either Bush or Dukakis, what is the posterior probability that there was a shift toward Bush between the two surveys?
\end{enumerate}

\bigskip

\item Suppose we have $n$ observations that follow a Normal distribution with a common mean $\mu$ and variance $\sigma^2$.  Also, suppose that we know the mean of the data, but want to learn about the variance of the data.  Find the posterior distribution of the variance $\sigma^2$ given an Inverse-gamma prior.  Specifically, find $p(\sigma^2 | \mathbf{y})$ given
\begin{eqnarray*}
Y_i &\sim& \text{N}(\mu, \sigma^2) \\
\sigma^2 &\sim& \text{Inv-gamma}(\alpha, \beta)
\end{eqnarray*}
\end{enumerate}

\end{document}





