\documentclass[10pt,a4paper]{article}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fullpage}
\usepackage{color}

\author{Patrick Lam}

\parindent=0in
\newcommand{\red}{\color{red}}
\newcommand{\black}{\color{black}}
\begin{document}

\begin{center}
\begin{Large}Week 5 Problems\end{Large}
\end{center}
\bigskip
\begin{enumerate}

\item Load the {\tt turnout} dataset from the {\tt Zelig} library.  Implement a Bayesian logistic regression of {\tt vote} on {\tt age} and {\tt income} using a random walk Metropolis-Hasting algorithm with a diffuse multivariate Normal prior.  Visually check for convergence.

\begin{itemize}
\item First, use a multivariate Normal jumping distribution to draw all the parameters at the same time.  Keep track of your acceptance rate.  Note any problems that you encountered.
\item Next, draw each $\beta$ separately with a univariate Normal jumping distribution given your draws of the other $\beta$s.  That is, draw in the following order:
\begin{enumerate}
\item[1.] $\beta_0^{(t)} | \beta_1^{(t-1)}, \beta_2^{(t-1)}$
\item[2.] $\beta_1^{(t)} | \beta_0^{(t)}, \beta_2^{(t-1)}$
\item[3.] $\beta_2^{(t)} | \beta_0^{(t)}, \beta_1^{(t)}$
\end{enumerate}
You may tune your individual jumping distributions differently (by setting different variances) to better control acceptance rates.
\end{itemize}

Compare your results to the MLE.  You may also check your results using {\tt MCMClogit()} in {\tt MCMCpack}.    
\end{enumerate}

\end{document}





